Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-12 ms.tex
Some minor changes; spellign etc

\newcommand{\vdag}{(v)^\dagger}
\newcommand{\myemail}{nkarlsson@physics.umn.edu}
\newcommand{\expval}[1]{\left\langle #1 \right\rangle}
\newcommand{\RA}[3]{#1$^{\mathrm{h}}$#2$^{\mathrm{m}}$#3$^{\mathrm{s}}$}
\newcommand{\Dec}[3]{#1$^{\circ}$#2\arcmin#3\arcsec}
\newcommand{\rmn}{\mathrm}
\newcommand{\CR}{\mathrm{CR}}
\newcommand{\dps}{\displaystyle}
\newcommand{\eps}{\varepsilon}
\newcommand{\bra}{\langle}
\newcommand{\ket}{\rangle}

\shorttitle{Coma Cluster Observations with VERITAS}
\shortauthors{VERITAS collaboration}

@@

\section{Introduction}
Clusters of galaxies are the largest virialized objects in the Universe, with typical diameters of
a few Mpc and masses on the order of $10^{14}$ to $10^{15} M_{\odot}$. According to the currently
favored hierarchical model of cosmic structure formation, larger objects formed through successive
mergers of smaller objects with galaxy clusters sitting a top this mass hierarchy
\citep[see][for a review]{article:Voit:2005}. Most of the mass of a cluster is in form of dark
matter (DM), as indicated by galaxy dynamics and gravitational lensing
\citep{article:DiaferioSchindlerDolag:2008}. Baryonic gas making up the intra-cluster medium (ICM)
contributes about 15\% of the total cluster mass and individual galaxies account for the remainder
(about 5\%). The ICM gas mass also comprises a significant fraction of the observable (baryonic)
matter in the Universe.

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung radiation in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated to these
temperatures primarily through collisionless structure-formation shocks that form as a result of the
hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two competing theories seeking to explain the radio halo
emission feature the ``hadronic model'' where the radio-emitting electrons are produced in inelastic
collisions of CR ions with thermal gas protons from the ICM \citep[][and references
therein]{article:EnsslinPfrommerMiniatiSubramanian:2011} and the ``re-acceleration model'' where a
CR electron population within the ICM at energies around 100 MeV might be sufficiently long lived so
that during states of powerful ICM turbulence, e.g., after a cluster merger, non-resonant
interactions of these CR electrons with plasma waves might be efficient enough to accelerate them to
high enough energies ($\sim 10$~GeV) so that they emit observable radio emission \citep[][and
references therein]{2010arXiv1008.0184B}.  Recent observations of possible non-thermal emission from
clusters in the the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays
\citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may
provide further indications of relativistic particle populations in clusters, although the
interpretation of these observations as non-thermal diffuse emission have been disputed on the basis
of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
  article:Ajello_etal:2010, article:Wik_etal:2009}.

Galaxy clusters have for many years been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic cosmic-ray (CR) hadrons,
in the form of protons and heavy ions, in the ICM is to be expected. The main energy-loss
mechanism for CR hadrons at high energies is pion production through the interaction of CRs with
nuclei in the ICM. Pions are short lived and decay; the charged pions to electrons and positrons,
\footnote{A charged pion decays to a muon or antimuon, which subsequently decay to an electron or
positron as well as neutrinos.} and the neutral pion to gamma rays. Due to the low density of the
ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large scales of a cluster and the volume
filling magnetic fields in the ICM, CR hadrons will be confined in the cluster on timescales
comparable and longer than the Hubble time \citep[][]{article:Volk_etal:1996,
article:Berezinsky_etal:1997} allowing them to accumulate over time. Depending on the CR pressure,
quantified by the CR-to-thermal pressure fraction
$X_\CR=\expval{P_{\CR}}/\expval{P_{\mathrm{th}}}$, an observable flux of high-energy gamma rays
can be expected.

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the
temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster
due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the
thermal electron pressure. This could then severely jeopardize the use of clusters to determine
cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy
clusters yields an upper limit of 20--30 \% of non-thermal pressure (that can be composed of CRs,
magnetic fields or turbulence) relative to the thermal gas pressure
\citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially
resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters
detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ --
within which the mean density is 500 times the critical density -- suggesting again a substantial
non-thermal pressure contribution on large scales \citep{article:Mahdavi_etal:2008}. High-energy
gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that
is most sensitive to the cluster core region. However, it assumes that the CR component is fully
mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal
ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux
observable with current generation gamma-ray telescopes rendering this technique at least as
sensitive as the dynamical and hydrostatic method (which are however more general in that they are
sensitive to any non-thermal pressure component).

@@

While several observations of clusters of galaxies have been made with satellite-borne and
ground-based telescopes, a detection of gamma-ray emission from a cluster has yet to be made.
Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the Fermi
Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on the
gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for Fermi observations) for several
galaxy clusters in the 100 MeV to GeV band. Upper limits on the VHE gamma-ray flux from a few
select clusters, including the Coma cluster, have been provided by observations with ground-based
imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Perkins_etal:2006,
inproc:Perkins_etal:2008, article:Aharonian_etal:2009a, article:Aleksic_etal:2010}.

@@

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter
imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Holder_etal:2006} located at an
altitude of 1268 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona
(31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes
is equipped with a 499 pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in
the spring of 2007, is designed to detect gamma-ray emission from astrophysical objects in the
energy range from 100 GeV to more than 30 TeV. The energy resolution is about 15\% and the angular
resolution (68\% containment) is about 0.1$^{\circ}$ per event at 1 TeV. The sensitivity of the
array allows for detection of a point source with a flux of 1\% of the steady Crab Nebula flux
above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The integral flux sensitivity
above 300 GeV was improved by about 30\% with the move of one telescope in the summer of 2009. The
move had no impact on the this data set.}

@@

Gamma-ray like events are separated from the CR background by imposing cuts on scaled parameters
\citep{article:Aharonian_etal:1997, article:Krawczynski_etal:2006} calculated from a parametrized
moment analysis of the shower images \citep{inproc:Hillas:1985}. The gamma-hadron separation cuts
used in this analysis were optimized a priori for a weak point source (3\% Crab Nebula flux level)
and a photon index of -2.4 using data taken on the Crab Nebula during the same epoch and are termed
\emph{standard} cuts. Because the VHE gamma-ray spectrum for the Coma cluster is expected to be
hard, photon index of about -2.3, these standard cuts are suitable for the analysis of the Coma
cluster data set.

@@

\section{VERITAS Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as
illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$
distribution is a radial comparison of the ON-source event distribution with the estimated
background. It extends out to a radius of $0.42^{\circ}$ to cover all regions of interest
considered in this work. A 99\% confidence level upper limit is calculated for each region of
interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian
background. A lower bound of zero is imposed on the signal rate to handle the negative excesses
seen for several of the regions of interest. Figure \ref{fig:sigdist} shows that the VERITAS data
is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent
of unity.

@@

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff on the CR distribution
function, i.e., we adopt $q=0$. Since a priori, the CR spectral index is unconstrained, we vary it
within plausible ranges of $2.1<\alpha_{\rmn{p}}<2.5$. This spectral range is compatible with the
radio spectral index of the giant radio halo of Coma after accounting for the spectral steepening at
frequencies $\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich effect
\citep{article:PfrommerEnsslin:2004b}\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
  responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
  times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.}.  To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

Table \ref{table:constraints} shows the resulting constraints on the relative CR pressure, $X_{\CR}
= \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius, $R_\rmn{vir}=2.2$~Mpc,
that we define as the radius of a sphere enclosing a mean density that is 200 times the critical
density of the Universe. The VERITAS flux upper limits strongly depend on $\alpha$ due to the
comparably large lever arm from GeV energies (that dominate the CR pressure) to TeV-energies. The
limits within 0.2$^{\circ}$ are most constraining yielding limits on $X_\CR$ between 0.048 and 0.43
(for $\alpha$ varying between 2.1 and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close
to the spectral index predicted by the simulations of \citet{article:PinzkePfrommer:2010} around 220
GeV).  The {\em Fermi} limits depend only weakly on $\alpha$ and are most constraining for an
aperture of 0.4$^{\circ}$.  The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$ and only
slightly worse for $\alpha\simeq 2.1$ yielding $X_\CR<0.025$.

@@

We complement the simplified analytical analysis with a more realistic and predictive approach
derived from cosmological hydrodynamical simulations. We adopt the universal spectral and spatial
gamma-ray model developed by \citet{article:PinzkePfrommer:2010} to estimate the emission from
decaying neutral pions that dominates over the IC emission from primary and secondary electrons
above 100 MeV in clusters.  Given a density profile as, e.g., inferred by cosmological simulations
or X-ray observations, the analytic approach models the CR distribution and the associated radiative
emission processes from radio to the gamma-ray band. This formalism was derived from high-resolution
simulations of clusters of galaxies that included radiative hydrodynamics, star formation, supernova
feedback, and followed CR physics by tracing the most important injection and loss processes
self-consistently while accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007,
  article:JubelgasSpringelEnsslinPfrommer:2008}. The results are in line with earlier numerical
results on some of the overall characteristics of the CR distribution and the associated radiative
emission processes \citep{article:DolagEnsslin, article:MiniatiRyuKangJones:2001,
  article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008}.

@@

We adopt the electron density profile as in Sect.~\ref{sec:simple} and model the temperature profile
of Coma with a constant central temperature of $kT= 8.25$~keV and a characteristic decline toward
the cluster periphery in accordance with a fit to the universal profile derived from cosmological
cluster simulations \citep{article:PinzkePfrommer:2010, article:Pfrommer_etal:2007} that also
follows the behavior of a nearby sample of deep {\em Chandra} cluster data
\citep{2005ApJ...628..655V}. This enables us to adopt the spatial and spectral distribution of CRs
according to the model by \citet{article:PinzkePfrommer:2010} that neglects the contribution of
cluster galaxies.

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted lines). Additionally shown are integrals of the differential spectrum for
finited energy intervals across the angular apertures tested in this study (solid lines). These
modeled fluxes a compared to {\em Fermi} and VERITAS flux upper limits of the same energy intervals,
summarized in Table \ref{table:constraints}.  The gamma-ray flux limit of {\em Fermi} in the energy
interval of 1--3 GeV ($<0.4^\circ$) is most constraining but still above the model predictions that
assume an optimistic value of the shock acceleration efficiency and CR transport parameters as laid
out above.  In our further analysis, we use the most constraining {\em Fermi} flux limits in the
energy interval of 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

@@

The expected gamma-ray flux due to self-annihilations of WIMPs is given by 
\begin{equation}
\frac{d\Phi_{\gamma}(\Delta\Omega,E)}{dE}=
\frac{\expval{\sigma v}}{8\pi m_{\chi}^{2}}\frac{dN(E,m_{\chi})}{dE} J(\Delta\Omega),
\label{eqn:WIMPflux}
\end{equation}
where $\expval{\sigma v}$ is the thermally averaged product of the total self-annihilation cross
section and the WIMP velocity, $m_{\chi}$ is the WIMP mass, $dN(E,m_{\chi})/dE$ is the differential
gamma-ray yield per annihilation, $\Delta\Omega$ is the observed solid angle, and $J$ is the
astrophysical factor - a dimensionless factor which determines the DM annihilation rate and depends
on the DM distribution.

Given the upper limit on the observed gamma-ray rate, $R_{\gamma}(95\%\ \mathrm{CL})$, we can place
constraints on the WIMP parameter space $(m_{\chi}, \expval{\sigma v})$. From Eq.
(\ref{eqn:WIMPflux}) we get
\begin{equation}
\expval{\sigma v}(95\%\ \mathrm{CL}) > 
R_{\gamma}(95\%\ \mathrm{CL}) \frac{8\pi m_{\chi}^{2}}{J(\Delta\Omega)}
\frac{1}{\int^{m_{\chi}} dE A_{\mathrm{eff}}dN_{\gamma}/dE}
\end{equation}
The astrophysical factor $J$ can be expressed as
\begin{equation}
J(\Delta\Omega)=\left(\frac{1}{\rho_{c}^{2}R_{H}}\right)
\int_{\Delta\Omega}d\Omega\int d\lambda \rho^{2}(\lambda,\Omega),
\end{equation}
where $\lambda$ represents the line-of-sight. Because the self-annihilation of a WIMP is a two-body
process, $J(\Delta\Omega)$ is proportional to the square of the DM density $\rho$. Here,
$J(\Delta\Omega)$ has been normalized to the square of the critical density, $\rho_{c}$, and the
Hubble radius, $R_{H}$.

In this work we have adopted the DM density profile by Navarro-Frenk-White
\citep[NFW;][]{article:NavarroFrenkWhite:1997},
\begin{equation}
\rho_{\chi}(r)=\rho_{s}\left(\frac{r}{r_{s}}\right)^{-1}\left(1+\frac{r}{r_{s}}\right)^{-2},
\end{equation} 
which parametrizes the distribution of DM by the scale radius $r_{s}$ and the scale density
$\rho_{s}$. [how where these calculated?]

%When the size of the integration region is increased, the limits on $\expval{\sigma v}$ result
%from a competition between the gain in the astrophysical factor $\expval{J}$ and the integrated
%background. For integration regions larger than 0.2$^{\circ}$ in radius, the astrophysical factor
%no longer compensate for the increased count of background events and the signal-to-noise
%deteriorates.

>>>>

\newcommand{\vdag}{(v)^\dagger}
\newcommand{\myemail}{nkarlsson@physics.umn.edu}
\newcommand{\expval}[1]{\left\langle #1 \right\rangle}
\newcommand{\RA}[3]{#1$^{\mathrm{h}}$#2$^{\mathrm{m}}$#3$^{\mathrm{s}}$}
\newcommand{\Dec}[3]{#1$^{\circ}$#2\arcmin#3\arcsec}
\newcommand{\rmn}{\mathrm}
\newcommand{\CR}{\mathrm{CR}}
\newcommand{\dps}{\displaystyle}
\newcommand{\eps}{\varepsilon}
\newcommand{\bra}{\langle}
\newcommand{\ket}{\rangle}
\shorttitle{Coma Cluster Observations with VERITAS}
\shortauthors{VERITAS collaboration}

@@

\section{Introduction}
Clusters of galaxies are the largest virialized objects in the Universe, with typical diameters of
a few Mpc and masses on the order of $10^{14}$ to $10^{15} M_{\odot}$. According to the currently
favored hierarchical model of cosmic structure formation, larger objects formed through successive
mergers of smaller objects with galaxy clusters sitting on top of this mass hierarchy
\citep[see][for a review]{article:Voit:2005}. Most of the mass of a cluster is in form of dark
matter (DM), as indicated by galaxy dynamics and gravitational lensing
\citep{article:DiaferioSchindlerDolag:2008}. Baryonic gas making up the intra-cluster medium (ICM)
contributes about 15\% of the total cluster mass and individual galaxies account for the remainder
(about 5\%). The ICM gas mass also comprises a significant fraction of the observable (baryonic)
matter in the Universe.

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung radiation in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated to these
temperatures primarily through collisionless structure-formation shocks that form as a result of the
hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two competing theories seeking to explain the radio halo
emission feature the ``hadronic model'' where the radio-emitting electrons are produced in inelastic
collisions of cosmic-ray (CR)  ions with thermal gas protons from the ICM \citep[][and references
therein]{article:EnsslinPfrommerMiniatiSubramanian:2011} and the ``re-acceleration model'' where a
CR electron population within the ICM at energies around 100 MeV might be sufficiently long lived so
that during states of powerful ICM turbulence, e.g., after a cluster merger, non-resonant
interactions of these CR electrons with plasma waves might be efficient enough to accelerate them to
high enough energies ($\sim 10$~GeV) for them to emit observable radio emission
\citep[][and references therein]{article:BrunettiLazarian:2010}. Recent observations of possible
non-thermal emission from clusters in the the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998}
and hard X-rays \citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004,
article:Eckert_etal:2007} may provide further indications of relativistic particle populations in clusters,
although the interpretation of these observations as non-thermal diffuse emission have been disputed on
the basis of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
article:Ajello_etal:2010, article:Wik_etal:2009}.

Galaxy clusters have for many years been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic CR hadrons,
in the form of protons and heavy ions, in the ICM is to be expected. The main energy-loss
mechanism for CR hadrons at high energies is pion production through the interaction of CRs with
nuclei in the ICM. Pions are short lived and decay; the charged pions to electrons and
positrons,\footnote{A charged pion decays to a muon or antimuon, which subsequently decay to an electron or positron as well as neutrinos.} and the neutral pion to gamma rays. Due to the low density of the
ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large scales of a cluster and the volume
filling magnetic fields in the ICM, CR hadrons will be confined in the cluster on timescales
comparable and longer than the Hubble time \citep[][]{article:Volk_etal:1996,
article:Berezinsky_etal:1997} allowing them to accumulate over time. Depending on the CR pressure,
quantified by the CR-to-thermal pressure fraction
$X_\CR=\expval{P_{\CR}}/\expval{P_{\mathrm{th}}}$, an observable flux of high-energy gamma rays
can be expected.

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the
temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster
due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the
thermal electron pressure. This could then severely jeopardize the use of clusters to determine
cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy
clusters yields an upper limit of 20--30 \% of non-thermal pressure (that can be composed of CRs,
magnetic fields or turbulence) relative to the thermal gas pressure
\citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially
resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters
detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ --
the radius within which the mean density is 500 times the critical density -- suggesting again a substantial
non-thermal pressure contribution on large scales \citep{article:Mahdavi_etal:2008}. High-energy
gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that
is most sensitive to the cluster core region. However, it assumes that the CR component is fully
mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal
ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux
observable with current generation gamma-ray telescopes rendering this technique at least as
sensitive as the dynamical and hydrostatic method (which are however more general in that they are
sensitive to any non-thermal pressure component).

@@

While several observations of clusters of galaxies have been made with satellite-borne and
ground-based gamma-ray telescopes, a detection of gamma-ray emission from a cluster has yet to be
made. Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the
Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on the
gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for Fermi observations) for several
galaxy clusters in the MeV to GeV band. Upper limits on the very-high-energy (VHE) gamma-ray flux
from a few select clusters, including the Coma cluster, have been provided by observations with
ground-based imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Perkins_etal:2006,
inproc:Perkins_etal:2008, article:Aharonian_etal:2009a, article:Aleksic_etal:2010}.

@@

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter
IACTs \citep{article:Holder_etal:2006} located at an altitude of 1268 m a.s.l. at the Fred Lawrence Whipple
Observatory in southern Arizona (31$^{\circ}$~40\arcmin~30\arcsec~N,
110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes is equipped with a 499 pixel camera
covering a 3.5$^{\circ}$ field of view. The array, completed in the spring of 2007, is designed to detect
gamma-ray emission from astrophysical objects in the energy range from 100 GeV to more than 30 TeV.
The energy resolution is about 15\% and the angular resolution (68\% containment) is about 0.1$^{\circ}$
per event at 1 TeV. The sensitivity of the array allows for detection of a point source with a flux of 1\% of
the steady Crab Nebula flux above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The
integral flux sensitivity above 300 GeV was improved by about 30\% with the move of one telescope in
the summer of 2009. The move had no impact on this data set.}

@@

Gamma-ray like events are separated from the CR background by imposing cuts on scaled parameters
\citep{article:Aharonian_etal:1997, article:Krawczynski_etal:2006} calculated from a parametrized
moment analysis of the shower images \citep{inproc:Hillas:1985}. The gamma-hadron separation cuts
used in this analysis were optimized a priori for a weak point source (3\% Crab Nebula flux level)
and a photon index of 2.4 using data taken on the Crab Nebula during the same epoch. Because the
VHE gamma-ray spectrum for the Coma cluster is expected to be hard, photon index of about 2.3,
these cuts are suitable for the analysis of the Coma cluster data set.

@@

\subsection{VERITAS Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as
illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$
distribution is a radial comparison of the ON-source event distribution with the estimated
background. It extends out to a radius of $0.42^{\circ}$ to cover all regions of interest
considered in this work. A 99\% confidence level upper limit is calculated for each region of
interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian
background. A lower bound of zero is imposed on the signal rate to handle the negative excesses
seen for several of the regions of interest. Figure \ref{fig:sigdist} shows that the VERITAS data
is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent
of unity.

@@

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff on the CR distribution
function, i.e., we adopt $q=0$. Since a priori, the CR spectral index is unconstrained, we vary it
within plausible ranges of $2.1<\alpha_{\rmn{p}}<2.5$. This spectral range is compatible with the
radio spectral index of the giant radio halo of the Coma cluster after accounting for the spectral
steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich effect
\citep{article:PfrommerEnsslin:2004b}.\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.} To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

Table \ref{table:constraints_simple} shows the resulting constraints on the relative CR pressure, $X_{\CR}
= \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius, $R_\rmn{vir}=2.2$~Mpc,
that we define as the radius of a sphere enclosing a mean density that is 200 times the critical
density of the Universe. The VERITAS flux upper limits strongly depend on $\alpha$ due to the
comparably large lever arm from GeV energies (that dominate the CR pressure) to TeV-energies. The
limits within 0.2$^{\circ}$ are the most constraining, yielding limits on $X_\CR$ between 0.048 and 0.43
(for $\alpha$ varying between 2.1 and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close
to the spectral index predicted by the simulations of \citet{article:PinzkePfrommer:2010} around 220
GeV).  The {\em Fermi} limits depend only weakly on $\alpha$ and are most constraining for an
aperture of 0.4$^{\circ}$.  The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$ and only
slightly worse for $\alpha\simeq 2.1$ yielding $X_\CR<0.025$.

@@

We complement the simplified analytical analysis with a more realistic and predictive approach
derived from cosmological hydrodynamical simulations. We adopt the universal spectral and spatial
gamma-ray model developed by \citet{article:PinzkePfrommer:2010} to estimate the emission from
decaying neutral pions that dominates over the inverse Compton (IC) emission from primary and
secondary electrons above 100 MeV in clusters.  Given a density profile as, e.g., inferred by
cosmological simulations or X-ray observations, the analytic approach models the CR distribution and
the associated radiative emission processes from radio to the gamma-ray band. This formalism was
derived from high-resolution simulations of clusters of galaxies that included radiative hydrodynamics,
star formation, supernova feedback, and followed CR physics by tracing the most important injection
and loss processes self-consistently while accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007,
  article:JubelgasSpringelEnsslinPfrommer:2008}. The results are in line with earlier numerical
results on some of the overall characteristics of the CR distribution and the associated radiative
emission processes \citep{article:DolagEnsslin:2000, article:MiniatiRyuKangJones:2001,
  article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008}.

@@

We adopt the electron density profile as in \S \ref{sec:simple} and model the temperature profile
of the Coma cluster with a constant central temperature of $kT= 8.25$~keV and a characteristic
decline toward the cluster periphery in accordance with a fit to the universal profile derived from
cosmological cluster simulations \citep{article:PinzkePfrommer:2010, article:Pfrommer_etal:2007} that
also follows the behavior of a nearby sample of deep {\em Chandra} cluster data
\citep{article:Vikhlinin_etal:2005}. This enables us to adopt the spatial and spectral distribution
of CRs according to the model by \citet{article:PinzkePfrommer:2010} that neglects the contribution
of cluster galaxies.

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted lines). Additionally shown are integrals of the differential spectrum for
finited energy intervals across the angular apertures tested in this study (solid lines). These
model fluxes are compared to {\em Fermi} and VERITAS flux upper limits of the same energy intervals,
summarized in Table \ref{table:constraints}.  The gamma-ray flux limit of {\em Fermi} in the energy
interval of 1--3 GeV ($<0.4^\circ$) is most constraining but still above the model predictions that
assume an optimistic value of the shock acceleration efficiency and CR transport parameters as laid
out above.  In our further analysis, we use the most constraining {\em Fermi} flux limits in the
energy interval of 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

@@

The expected gamma-ray flux due to self-annihilations of WIMPs is given by 
\begin{equation}
\frac{d\Phi_{\gamma}(\Delta\Omega,E)}{dE}=
\frac{\expval{\sigma v}}{8\pi m_{\chi}^{2}}\frac{dN(E,m_{\chi})}{dE} J(\Delta\Omega),
\label{eqn:WIMPflux}
\end{equation}
where $\expval{\sigma v}$ is the thermally averaged product of the total self-annihilation cross
section and the WIMP velocity, $m_{\chi}$ is the WIMP mass, $dN(E,m_{\chi})/dE$ is the differential
gamma-ray yield per annihilation,\footnote{In this work, we have calculated the differential
gamma-ray yield per annihilation using the Pythia Monte Carlo simulator.} $\Delta\Omega$ is the
observed solid angle, and $J$ is the astrophysical factor - a dimensionless factor which determines
the DM annihilation rate and depends on the DM distribution.

Given the upper limit on the observed gamma-ray rate,
$R_{\gamma}(95\%\ \mathrm{CL}) = N_{\gamma}(95\%\ \mathrm{CL}) / T_{\mathrm{obs}}$, we can place
constraints on the WIMP parameter space $(m_{\chi}, \expval{\sigma v})$. From Eq.
(\ref{eqn:WIMPflux}) we get
\begin{equation}
\expval{\sigma v}(95\%\ \mathrm{CL}) > 
R_{\gamma}(95\%\ \mathrm{CL}) \frac{8\pi m_{\chi}^{2}}{J(\Delta\Omega)}
\frac{1}{\int^{m_{\chi}} dE A_{\mathrm{eff}}dN_{\gamma}/dE}
\end{equation}
Because the self-annihilation of a WIMP is a two-body process, the astrophysical factor
$J(\Delta\Omega)$ can be expressed as the line-of-sight integral of the DM density squared
\begin{equation}
J(\Delta\Omega)=\left(\frac{1}{\rho_{c}^{2}R_{H}}\right)
\int_{\Delta\Omega}d\Omega\int d\lambda \rho^{2}(\lambda,\Omega),
\end{equation}
where $\lambda$ represents the line-of-sight. Here,
$J(\Delta\Omega)$ has been normalized to the square of the critical density, $\rho_{c}$, and the
Hubble radius, $R_{H}$. In this work we have adopted the DM density profile by
Navarro-Frenk-White \citep[NFW;][]{article:NavarroFrenkWhite:1997},
\begin{equation}
\rho_{\chi}(r)=\rho_{s}\left(\frac{r}{r_{s}}\right)^{-1}\left(1+\frac{r}{r_{s}}\right)^{-2},
\end{equation} 
which parametrizes the distribution of DM by the scale radius $r_{s}$ and the scale density
$\rho_{s}$. Using weak-lensing measurements of the virial mass in the Coma cluster
\citep{article:Gavazzi_etal:2009} we obtained the values $r_{s}=0.654$ Mpc and
$\rho_{s}=4.4\times 10^{4}$ M$_{\odot}$/Mpc$^{3}$. Table \ref{table:astrofactor} lists the
astrophysical factor calculated for the different VERITAS appertures considered in this work.
It also lists the astrophysical factor calculated for the background region, which has been used
to estimate the gamma-ray contamination due to DM in the background region.
